---
title: "Using sRACIPE"
author: "Vivek Kohar"
date: "10/4/2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document shows how to use sRACIPE to simulate any network topology and analyze the generated data. For simplicity, we will use the toggle switch with mutual inhibition and self activation of both genes as a test case.

## Load the network (topology)
As a first step set the directory to the desired folder where you want to keep the input topology file and the results. In the working directory ("man" in this example), create a folder "inputs" and place the topology (the network) file in that folder. The typical format of the topology file is a 3 column file where the first column is name of the source gene, second column is name of the target gene and last column is the interaction type (1 - activation, 2-inhibition). The first line should contain the header (Source Target Interaction). 

```{r}
rm(list = ls())
library(sRACIPE)
working_directory <- getwd()
topology_file <- "inputs/test.tpo"
topology <- sRACIPE_load_topology(topology_file = topology_file)
topology

```

## Initialize the simuation parameters

A configuration file containing the parameters for network simulation can be placed in the inputs folder. If the configuration file is not available, default arguments are used. Alternatively, you can specify the individual parameters when calling the functions.
```{r}
config_file = "inputs/sRACIPE.cfg"
configuration <- sRACIPE_load_configuration(config_file = config_file)
configuration
```

## Deterministic simulations using fourth order Runge-Kutta method.

Now we will show deterministic simulations of this network using the fourth order Runge-Kutta integration method. Calling this function will generate four output files in the "results" folder in the same directory. The files contain (a) the names of the genes in the network, (b) the parameters for all the models, (c) the initial conditions used for each model (d) the final gene expressions for all the genes.
```{r}
output_file <- sRACIPE_RK_deterministic(topology_file = topology_file, NUM_MODELS = 2000, INITIAL_CONDITIONS = 1)

```

## Plot the results in a pdf file in the results folder. There are three plots, (a) Heatmap of gene expressions on the first and second PC components, (b) Hierarchical clustering of gene expressions, (c)  t-SNE heatmap of gene expressions   
```{r}
plot_data_deterministic(output_file, plot_filename = topology$filename)
```

## Plot the network in a html file in the results  folder

```{r}
plot_network(topology)
```


## Stochastic simulations

##  Stochastic simulations without annealing (for multiple noise levels)
Now we will perform stochastic simulations of this network using the Euler-Maruyama integration method. 

Calling "sRACIPE_stochastic"" function will generate four output files in the "results" folder in the same directory. The files contain (a) the names of the genes in the network, (b) the parameters for all the models, (c) the initial conditions used for each model (d) the final gene expressions for all the genes. Notice that the output files now have NOISE_LEVELS*number_gene columns.

```{r}
output_file_stochastic_NoAnnealing <- sRACIPE_stochastic(topology_file = topology_file, MAX_NOISE = 30, NOISE_LEVELS = 5, NUM_MODELS = 2000, INITIAL_CONDITIONS = 1, ANNEAL=F, NOISE_SCALING_FACTOR=0.6, SIM_TIME = 100, STEP_SIZE = 0.02)

```

## Plot the stochastic results in a pdf file in the results folder

This will plot the heatmap of gene expressions on the first two principal components. PCs are calculated using the determinstic data (D=0) and stochastic data (D > 0) are projected on the same PCs

```{r}
plot_data_stochastic(output_file_stochastic_NoAnnealing, plot_filename=topology$filename, topology_df=topology, config = configuration, bin_count=40)
```

##  Stochastic simulations with annealing.

We can perform stochastic simulations with annealing in similar way. Typically large number of noise levels (~30) is needed for better convergence


```{r}
output_file_stochastic_Annealing <- sRACIPE_stochastic(topology_file = topology_file, MAX_NOISE = 30, NOISE_LEVELS = 30, NUM_MODELS = 2000, INITIAL_CONDITIONS = 1, ANNEAL=F, NOISE_SCALING_FACTOR=0.6, SIM_TIME = 100, STEP_SIZE = 0.02)

```

## Plot the stochastic results with annealing in a pdf file in the results folder

This will plot the heatmap of gene expressions on the first two principal components. PCs are calculated using the determinstic data (D=0) and stochastic data (D > 0) are projected on the same PCs

```{r}
plot_data_stochastic(output_file_stochastic_Annealing, plot_filename=topology$filename, topology_df=topology, config = configuration, bin_count=40)
```

