---
title: "A systems biology tool for gene regulatory circuit simulation"
author:
- name: Vivek Kohar
  affiliation: The Jackson Laboratory, Bar Harbor, Maine, USA
- name: Aidan Tillman
  affiliation: Department of Bioengineering, Northeastern University, Boston, Massachusetts, USA
output:
  BiocStyle::html_document:
    toc_float: true
  BiocStyle::pdf_document: default
package: sRACIPE
vignette: |
  %\VignetteIndexEntry{A systems biology tool for gene regulatory circuit simulation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 7, fig.height = 5)
```

# Introduction

This document illustrates the use of sRACIPE to simulate any 
circuit or network or topology (used interchangeably) 
and analyze the generated data. sRACIPE implements a 
randomization-based method for gene circuit modeling. 
It allows us to study the effect of both the gene expression noise 
and the parametric variation on any gene regulatory circuit (GRC) 
using only its topology, and simulates an ensemble of models 
with random kinetic parameters at multiple noise levels. 
Statistical analysis of the generated gene expressions 
reveals the basin of attraction and stability of various 
phenotypic states and their changes associated with 
intrinsic and extrinsic noises. sRACIPE provides a 
holistic picture to evaluate the effects of both the 
stochastic nature of cellular processes and the parametric variation. 

## Installation
1. Download the package from Bioconductor.
```{r getPackage, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
install.packages("BiocManager")
BiocManager::install("sRACIPE")
```

Or install the development version of the package.
```{r, eval = FALSE}
BiocManager::install(“lusystemsbio/sRACIPE”)
```

2. Load the package into R session.

```{r Load, message=FALSE}
suppressWarnings(suppressPackageStartupMessages(library(sRACIPE)))
```

For simplicity, we will start with the toggle switch 
with mutual inhibition and 
self activation of both genes as a test case. Additionally,
to keep the simulations fast, we will use fewer models, 
small integration time and longer integration step size.
We recommend using the default settings for most 
parameters for actual simulations.

# Load the Circuit
One can use the topology stored in a text file or loaded 
as a dataframe. The typical format of the topology 
file is a 3 column file where the first column is name of the source gene, 
second column is name of the target gene and last column is 
the interaction type (1 - TF activation, 2- TF inhibition, 3- activation by 
inhibiting degradation, 4- inhibition by activating degradation, 5- signalling
activation, 6- signaling inhibition). 
The first line should contain the header (Source Target Interaction). 
We will work with a demo circuit from the package.

```{r eval=TRUE}
suppressWarnings(suppressPackageStartupMessages(library(sRACIPE)))

# Load a demo circuit
data("demoCircuit")
demoCircuit

```

# Simulate the circuit
We will use a reduced number of models (using numModels) for demonstration. 
By default, deterministic simulations are run with convergence testing, so we
alter the numConvergenceIter parameter to change how long a simulation lasts.
The simulations will return a RacipeSE object. Convergence data is stored in 
the colData of the RacipseSE object and can be retrieved using the 
sracipeConverge method.


```{r}

rSet <- sRACIPE::sracipeSimulate(circuit = demoCircuit, numModels = 20,
                             plots = FALSE, integrateStepSize = 0.1, 
                             numConvergenceIter = 5)
cd <- sracipeConverge(rSet)

```

# Plotting the simulated data

We can plot the simulated data using the sracipePlotData 
function or using plots=TRUE
in sracipeSimulate. The data can be normalized before plotting. 
Otherwise it will be normalized by the plotting function. By default, two 
clusters are identified and models colored according to hierarchical clustering.

```{r eval=TRUE}
rSet <- sRACIPE::sracipeNormalize(rSet)
rSet <- sRACIPE::sracipePlotData(rSet, plotToFile = FALSE)

```

# Multistability analysis

One workflow in deterministic analysis is to simulate multiple initial 
conditions for each model and collect the unique steady states of each model
for multistability analysis. This is done in sRACIPE by the sracipeUniqueStates
method, which takes in a RacipeSE object and returns a list of the steady states
for each model. Since the method only considers converged expressions, the 
following code may not produce anything, so consider trying it with larger 
simulations

```{r eval=FALSE}
data("demoCircuit")

rSet <- SRACIPE::sracipeSimulate(circuit = demoCircuit, numModels = 10,
                             plots = FALSE, integrateStepSize = 0.1, 
                             numConvergenceIter = 5, nIC = 2)
stateLst <- sracipeUniqueStates(rSet)

```


# Knockdown Analysis
The simulations can be used to perform in-silico perturbation analysis. 
For example, here we will limit a gene's production rate to mimic its knockdown
and show how that changes the relative proportion of models in different 
clusters.

```{r eval=TRUE}
data("demoCircuit")
rSet <- sRACIPE::sracipeSimulate(circuit = demoCircuit, 
                             numModels = 50, plots = FALSE, 
                             integrateStepSize = 0.1, 
                             numConvergenceIter = 5)
kd <- sRACIPE::sracipeKnockDown(rSet, plotToFile = FALSE,
                                reduceProduction=50)
```

# Plot the network 

The network can be plotted in an interactive viewer 
or html file in the results  folder.

```{r eval=FALSE}
 sRACIPE::sracipePlotCircuit(rSet, plotToFile = FALSE, physics = TRUE)
```


# Limit Cycle Detection
For deterministic simulations, one can have the simulation check for limit 
cycles in the results, although the simulation won't do this by default. The
demoCircuit we were previously using can't produce oscillations, so we will have
to call another circuit, the repressilator, which is a loop of 3 genes 
inhibiting each other. Limit cycle results are stored in the metadata of the 
results, just like the convergence data. Note that with the small number of 
models simulated and low number of iterations, this code might not find any 
limit cycles in one try. Use a large number of models to guarantee limit cycle 
detection.

```{r eval=TRUE}
data("repressilator")
rSet <- sRACIPE::sracipeSimulate(circuit = repressilator, 
                             numModels = 30, plots = FALSE, 
                             integrateStepSize = 0.1, 
                             numConvergenceIter = 5,
                             limitcycles = TRUE)
```

# Gene Clamping Simulations

Clamping a gene is to keep it at a constant value for the entire simulation, 
regardless of interactions. This is done by supplying a data.frame to the 
geneClamping parameter in sracipeSimulate(), where the columns are genes and
the rows are models. If the user supplies only one value for each clamped gene,
then the gene is clamped to that value for all models.
```{r eval=TRUE}
rSet <- sracipeSimulate(circuit = demoCircuit,
                        numConvergenceIter = 5,
                        numModels = 10,
                        integrateStepSize = 0.1, 
                        geneClamping = data.frame(A = c(1))
                        )
```

#  Stochastic simulations 

One can perform stochastic simulations similarly by specifying additional 
parameters to the sracipeSimulate function, namely, nNoise (the number of noise
levels at which the stochastic simulations should be carried out), initialNoise 
(the starting noise level) and noiseScalingFactor (the multiplicative factor by
which noise should be reduced for multiple noise levels). 
For annealing, use anneal=TRUE alongwith the above mentioned parameters.
For simulations at one
noise level only, use nNoise = 1 and set initialNoise parameter to the specific 
noise. 
Since convergence testing is not possible for stochastic simulations, they are 
not run and we use the simulationTime parameter to modulate simulation length

Now the returned object will contain additional elements in the 
assays which correspond to the 
simulations at different noise levels 
(noise specified by the name of the element).

```{r eval=TRUE}
rSet <- sRACIPE::sracipeSimulate(circuit = demoCircuit, numModels = 20, 
                             initialNoise = 15, noiseScalingFactor = 0.1,
                             nNoise = 2,
                             plots = TRUE, plotToFile = FALSE, 
                             integrateStepSize = 0.1, 
                             simulationTime = 30)

```

Here, calling the sracipeSimulate function simulated the circuit at zero noise level
as well as the two (nNoise) other noise levels 15 (initialNoise), 1.5 
(initialNoise*noiseScalingFactor). The first three plots (hierarchical 
clustering heatmap, Umap, PCA) correspond to deterministic data and the last
two plots contain the data from stochastic simulations projected on the 
principal components of the deterministic simulations.

Note that the rSet now contains stochastic simulations as well as 
additional elements like umap, pca, assignedClusters. These are 
added when the data is plotted. As mentioned previously, 
the additional elements of the assays 
are named "15" and "1.5" which correspond to noise levels. 

For annealing simulations, one can set anneal=TRUE in the sracipeSimulate function.
With anneal=FALSE (constant noise), simulations at different noise levels are 
independent of each other. These are useful if one is primarily interested in 
the gene expressions at different noise levels and at zero noise 
(used for normalizing the data). With annealing, the steady state solutions
at higher noise levels are used as the intial conditions for lower noise levels 
such that each model converges to its most stable state when the noise is zero.


Using annealing, ideally the number of noise levels should be very large 
and noiseScalingFactor close to 1 as we want to reduce the noise very slowly. 
In practice, we found nNoise ~ 30 and 
initialNoise ~50 divided by sqrt(number_gene) 
as good starting values. Constant noise and annealing noise simulations pca 
plots can be used for better approximations of these parameters. 
The initialNoise should be such that there is a single cluster at this 
high noise level (essentially the gene expression values are random and circuit 
topology has little effect). Next, noiseScalingFactor  should be adjusted such 
that there are sufficient noise levels when this single cluster splits into 
multiple clusters observed in deterministic simulations. 

With annealing,
the models converge to their most stable steady state 
at zero noise. Thus, the number of models is more stable clusters
will increase and number in less 
stable clusters will decrease. Note that for non zero noise, the stable states
can be different from the stable states at zero noise. In our illustrative 
example shown abpve, the previous two stable states of a toggle circuit are no 
longer stable at high noise ("15") and instead the previously unstable high high
state is stable now. Briefly, noise can change the stable states and zero noise
simulations using annealing can  
gives us an idea about the relative stability of states 
when there are multiple stable states. 

Further, one can modify the parameters and/or initial conditions
and simulate the circuit with
modified parameters and/or initial conditions using the parameters 
genParams = FALSE and/or genIC = FALSE. 


```{r eval=TRUE}
rSet <- sRACIPE::sracipeSimulate(circuit = demoCircuit, numModels = 20, 
                             plots = FALSE, integrate = FALSE)
params <- sRACIPE::sracipeParams(rSet)
modifiedParams <- as.matrix(params) 
modifiedParams[,1] <- 0.1*modifiedParams[,1]
sRACIPE::sracipeParams(rSet) <- DataFrame(modifiedParams)
rSet <- sRACIPE::sracipeSimulate(rSet, plots = FALSE, genParams = FALSE)

```


# KnockOut Simulations

Knockout of a gene is implemented by changing the production rate and initial 
condition of the gene to zero. The knockOut parameter in the function 
sracipeSimulate can be used to perform these knockout simulations. If simulations 
are to be carried out for knockouts of different genes, the genes should be 
specified as a list where each list element will contain the names of the gene 
to be knocked out. For example, knockout = list("gene1", "gene2", 
c("gene3", "gene4"), "gene5") will knockout gene1, gene2, gene5 one by one
and knockout gene3 and gene4 simultaneously.
knockOut = "all", each gene is knocked out one by one and the results are 
returned as an element knockOutSimulations which, similar to 
stochasticSimulations, is a list of dataframes containing the gene expressions
obtained by knockout one or more genes. Enabling plots=TRUE will plot the 
results. As the expression of knockout gene is zero, we compute PCA with 
unperturbed genes for both the unperturbed simulations as well as the
perturbed simulations. So for each knockout, we have two plots containing the
scatter plot of unperturbed simulations and perturbed simulations on the PCs of
unperturbed simulations (excluding the gene to be perturbed).

# References
Kohar V, Lu M (2018). “Role of noise and parametric variation in the dynamics of gene regulatory circuits.” npj Systems Biology and Applications, 4, 40. https://www.nature.com/articles/s41540-018-0076-x.


# Session Information

```{r}
sessionInfo()
```

