---
title: "Simulating trajectories"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Simulating a circuit
Typically we use sRACIPE to simulate the steady state solutions but one can also
simulate the trajectories to study the effect of perturbations on the clusters.
Here we will show a pipeline to study the effect of perturbations on the clusters.

```{r }
# Load sRACIPE package
suppressWarnings(suppressPackageStartupMessages(library(sRACIPE)))
set.seed(123)
racipe <- RacipeSE() # Construct an empty RacipeSE object
data("EMT2") # Load a sample circuit
sracipeCircuit(racipe) <- EMT2
# Simulate the circuit
racipe <- sracipeSimulate(racipe, numModels = 100, plots = FALSE, genIC = TRUE, 
                         genParams = TRUE, integrate = TRUE, 
                         integrateStepSize = 0.05,
                         simulationTime = 50
                         )
# Calculate the mean and sd from the unnormalized data.
# These will be later used to normalize any new data.
simExp <- assay(racipe)
simExp <- log2(1+simExp)
tmpMeans <- rowMeans(simExp)
tmpSds <- apply(simExp,1,sd)

# Normalize the data
racipeNorm <- sracipeNormalize(racipe)
simExp <- assay(racipeNorm) # extract the simulated gene expression
pca <- prcomp(t(simExp)) # Plot the simulated data on PCA
## Define a function to plot pca datapoints
# Load the ggplot2 
suppressWarnings(suppressPackageStartupMessages(library(ggplot2)))

sracipePlotPcaPoints <- function(plotData, xmin=-2,xmax=2,ymin=-2,ymax=2){
  pcaData <- data.frame(x=plotData[,1],y=plotData[,2])
  require(ggplot2)
  return(ggplot2::ggplot(pcaData) +
      geom_point(aes(x = pcaData[,1], y =pcaData[,2])) +
      xlab(paste0("PC1(",100*summary(pca)$importance[2,1],"%)")) +
      ylab(paste0("PC2(",100*summary(pca)$importance[2,2],"%)")) +
      xlim(xmin,xmax) +
      ylim(ymin,ymax) +
      theme_bw() +
      NULL)
}

sracipePlotPcaPoints(pca$x,xmin=-6,xmax=6,ymin=-4,ymax=4)

```


## Selecting models belonging to a cluster
Now we can cluster the models and perturb models belonging to that
cluster. 

```{r }
# Define the distance function
distfun=function(x) as.dist((1-cor(t(x), method = "s"))/2)
distance <- distfun(t(simExp)) # Compute the distance
# Perform heirarchical clustering
clusters <- hclust(distance, method = "ward.D2" )
# Select the number of clusters
nClusters <- 3
# Obtain the cluster assignment of each model 
assignedClusters <- cutree(clusters, nClusters)

# Define a function to plot the heirarchical clustering data
sracipePlotHeatmap <- function(plotData, assignedClusters, col = NULL, 
                               col2 = NULL, 
                               ...) {
  if(is.null(col))  col <- c("#5E4FA2", "#4F61AA", "#4173B3", "#3386BC", "#4198B6",
               "#51ABAE", "#62BEA6", "#77C8A4", "#8ED1A4", "#A4DAA4",
               "#B8E2A1", "#CBEA9D", "#DEF199", "#EAF69F", "#F2FAAC",
               "#FAFDB8", "#FEFAB6", "#FEF0A5", "#FEE695", "#FDD985",
               "#FDC978", "#FDB96A", "#FCA75E", "#F99254", "#F67D4A",
               "#F26943", "#E85A47", "#DE4B4B", "#D33C4E", "#C1284A",
               "#AF1446", "#9E0142")
  if(is.null(col2))  col2 <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2",
          "#D55E00", "#CC79A7")

    clustNames <- unique(assignedClusters)
    nClusters <- length(clustNames)
    clustColors <- numeric(length(assignedClusters))
    for(tmp1 in seq_len(length(clustColors))){
      clustColors[tmp1] <- which(clustNames == assignedClusters[tmp1] )
    }
    clustColors <- col2[clustColors]
   names(clustColors) <- as.character(assignedClusters)
   require(gplots)
  gplots::heatmap.2(plotData, 
                    col=col, 
                    hclustfun = function(x) hclust(x,method = 'ward.D2'), 
       distfun=function(x) as.dist((1-cor(t(x), method = "s"))/2), 
                    trace="none",
  ColSideColors = clustColors, labCol = FALSE,
    xlab="Models",
    margins = c(2, 5)
    )
}
# Plot the data
sracipePlotHeatmap(simExp, assignedClusters = assignedClusters)
# Select the models belonging to first cluster
selectedModels <- which(assignedClusters ==1) 
# Plot the selected models on PC plot
tmp <- pca$x[selectedModels,]
sracipePlotPcaPoints(plotData = tmp,xmin=-6,xmax=6,ymin=-4,ymax=4)

```

## Simulating a subset of models
Now we select the models belonging to first cluster and decrease the ZEB1 
production rate.

```{r }
# Create a new racipe object with models from the selected cluster
subRacipe <- RacipeSE(racipe[,selectedModels])
subRacipe  
numModels <- dim(subRacipe)[2]
# modify the parameters
params <- sracipeParams(subRacipe)
params["G_ZEB1"] <- 0.01*params["G_ZEB1"]
sracipeParams(subRacipe) <- params

# Use the steady state solutions of the models as the initial conditin
ic <- assay(subRacipe,1)
sracipeIC(subRacipe) <- ic


# We will record the state at 0.05 intervals. For an integrateStepSize of 0.01,
# this implies that simulation results at every 5th time point are recorded.
subRacipeTS <- sracipeSimulate(subRacipe, numModels = numModels, 
                             plots = FALSE, genIC = FALSE, 
                         genParams = FALSE, integrate = TRUE, 
                         integrateStepSize = 0.01,
                         simulationTime = 10, printInterval = 0.05, printStart = 0
                         )
```

## Preprocessing time series data
```{r}
# Lets define a function to process the time series data.
# We should process all data uniformly to observe how changing a parameter
# affects the trajectories of the cluster.
# Use the mean and standard deviation from the original simulation to normalize.

processData <- function(racipe,counter,time){
  newExpData <- assay(racipe,counter) # data from different timepoints
  newExpData <- log2(1+newExpData) # Log transform
  newExpData <- sweep(newExpData, 1, tmpMeans, FUN = "-") # scale
  newExpData <- sweep(newExpData, 1, tmpSds, FUN = "/") # scale
  newPca <- (scale(t(newExpData), pca$center, pca$scale) %*%
             pca$rotation)
  return(data.frame(PC1=newPca[,1],PC2=newPca[,2], Time = time))
}
trajectoriesAll <- data.frame()
trajectoriesAll <- data.frame(processData(subRacipe,1,0))
models <- seq(1:numModels)
trajectoriesAll$Model <- models
nTimePoints <- length(assays(subRacipeTS))
for(i in 2:nTimePoints)
{
  exp <- processData(subRacipeTS,i,(i-1)*0.05)
  exp$Model <- models
  trajectoriesAll <- rbind(trajectoriesAll,exp)
}

p <- ggplot2::ggplot(trajectoriesAll) +
  xlab(paste0("PC1(",100*summary(pca)$importance[2,1],"%)")) +
  ylab(paste0("PC2(",100*summary(pca)$importance[2,2],"%)")) +
  geom_point(aes(x = PC1, y =PC2, color = Time), alpha = 0.5) +
 # ggtitle(" Time = 2000") +
  theme_bw() +
  theme(text = element_text(size=15))

p
```

## Plotting averaged values with errorbars
We can also observe how the mean and standard deviation of the models change
upon parameter perturbation. The graph shown below shows how the PC1 and PC2 change over time.
```{r }

pcaSimValues <- do.call(data.frame, aggregate(. ~ Time, trajectoriesAll[,1:3], function(x) c(mean = mean(x), sd = sd(x))))
p <- ggplot2::ggplot(pcaSimValues) +
  geom_point(aes(x = PC1.mean, y =PC2.mean, color = Time), alpha = 1) +
  geom_line(aes(x = PC1.mean, y =PC2.mean),size=3) +
  geom_errorbar(aes(x = PC1.mean, color = Time,ymin = (PC2.mean - PC2.sd),ymax = (PC2.mean + PC2.sd)), alpha = 0.1) +
  geom_errorbarh(aes( y =PC2.mean, color = Time,xmin = (PC1.mean - PC1.sd),xmax = (PC1.mean + PC1.sd)), alpha = 0.1) +
  xlab(paste0("PC1(",100*summary(pca)$importance[2,1],"%)")) +
  ylab(paste0("PC2(",100*summary(pca)$importance[2,2],"%)")) +
  theme_bw() +
  theme(text = element_text(size=20),
        legend.text = element_text(size=15),
        legend.spacing.y = unit(.10, 'cm'),
        legend.spacing.x = unit(.10, 'cm')
  ) 
p


```

## Creating a movie
One can also create a movie using `gganimate`.
```{r eval=FALSE}

library(gganimate)
library(transformr)

p <- ggplot2::ggplot(pcaSimValues[1:100,]) +
  geom_point(aes(x = PC1.mean, y =PC2.mean, color = Time)) +
  xlab(paste0("PC1(",100*summary(pca)$importance[2,1],"%)")) +
  ylab(paste0("PC2(",100*summary(pca)$importance[2,2],"%)")) +
  theme_bw() +
  transition_states(
    Time,
 wrap=FALSE
  ) +
  shadow_mark() +
  enter_fade() +
  exit_shrink() +
  ease_aes('sine-in-out')
#  NULL
animate(p, renderer = ffmpeg_renderer(),nframes = 199)
```
