---
title: "Single Model Simulations"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setting the circuit
This article shows how single model can be simulated using sRACIPE. 

```{r }

suppressWarnings(suppressPackageStartupMessages(library(sRACIPE)))
suppressWarnings(suppressPackageStartupMessages(library(ggplot2)))

racipe <- RacipeSE() # Construct an empty RacipeSE object
data("EMT1") # Load a sample circuit
sracipeCircuit(racipe) <- EMT1
racipe

```

## Generate a random parameter set

One can initialize the racipe object with experimentally determined parameters
or generate a random parameter set and then modify the some or all of the 
parameters. As we are interested in the trajectories of a single model, we set
`timeSeries` to `TRUE`.

```{r }
racipe <- sracipeSimulate(racipe, timeSeries = TRUE, plots = FALSE, genIC = TRUE, 
                         genParams = TRUE, integrate = FALSE
                       )

racipe

```
## Modify the parameters or initial conditions

Notice that the `colData` changed now and contains the randomly generated 
parameters and initial conditions. These can be accessed and modified using
`sracipeParams` and sracipeIC `accessors`.

```{r }
parameters <- sracipeParams(racipe) # Get the parameters
dim(parameters) 
parameters["G_Zeb1"]
parameters[,2]
parameters["G_Zeb1"] <- 10*parameters["G_Zeb1"] # modify parameters
parameters[,2] <- 50 # modify parameters
sracipeParams(racipe) <- parameters # change the parameters to modified values
racipe$G_Zeb1


ic <- sracipeIC(racipe) # Get the initial conditions
dim(ic) 
ic[1]

ic[1] <- 10 # modify initial condition
sracipeIC(racipe) <- ic # change the ic to modified values

```

## Simulate with modified values
  
```{r }
racipe <- sracipeSimulate(racipe, timeSeries = TRUE, plots = FALSE, genIC = FALSE, 
                         genParams = FALSE, integrate = TRUE,
                        simulationTime  = 20,
                       integrateStepSize = 0.05
                       )
# Function to plot data
sracipePlotTS <- function(plotData, ...){
  plotData <- t(plotData)
  sexprs <- stack(as.data.frame(plotData))
  colnames(sexprs) <- c("geneExp", "Gene")
  sexprs$time <- rep(as.numeric(rownames(plotData)), ncol(plotData))
  theme_set(theme_bw(base_size = 10))
  ggplot2::qplot(time, geneExp, data = sexprs, group = Gene, colour = Gene,
      geom = "line", ylab = "Gene Expression", xlab = "time" )
}
plotData <- sracipeGetTS(racipe)
p <- sracipePlotTS(plotData)
p + scale_y_log10() # Use log scale

```

## Parameter perturbation
The role of parameter perturbation can be studied by slowly changing one 
parameter while keeping the other paramters constant. Here we will simulate 
a large number of models with same parameters except for the parameter being 
perturbed. We start with a random initial condition each time to capture 
multiple attractors incase the system is multistable. 

```{r }
selectedParameter <- "G_Zeb1" # parameter to be perturbed
parMin <- 1 # Minimum value of parameter
parMax <- 100 # Maximum value of parameter
nModels <- 30 # Number of models. Parameter value will be uniformly sampled 
# nModels times from parMin to parMax.

parameters <- sracipeParams(racipe)
newParameters <- parameters[rep(seq_len(nrow(parameters)), nModels),]
tmpValue <- seq(from = as.numeric(parMin), 
                  to = as.numeric(parMax),
                  length.out = nModels)
newParameters[selectedParameter] <- tmpValue
    
circuit <- sracipeCircuit(racipe)
rs <- sracipeSimulate(circuit, genIC = TRUE, genParams = TRUE, integrate = FALSE,
                          numModels = nModels)
sracipeParams(rs) <- newParameters # Modify parameters etc

rs <- sracipeSimulate(rs, genIC = FALSE, genParams = FALSE, integrate = TRUE,
                          numModels = nModels, simulationTime = 100)

# Plot the steady state values
library(ggplot2)
sexprs <- assay(rs,1)
sexprs <- reshape2::melt(t(sexprs))
colnames(sexprs) <- c("bifurParameter","Gene","geneExp")
modParameter <- rep(tmpValue,times = dim(rs)[1])
sexprs$modParameter <- modParameter
theme_set(theme_bw(base_size = 10))
p <- ggplot2::ggplot(sexprs) +
        geom_point(aes(x=modParameter,y=geneExp,color = Gene)) +
        xlab((selectedParameter)) +
        ylab("Gene Expression") 
p + scale_y_log10() # Use log scale
   
```

