---
title: "Under the hood - design and implementation"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

We discuss the design of sRACIPE in this article. As our primary aim is to develop 
a robust framework for simulating gene expression profiles that can capture
the experimental observations, we utilized the well established data structures 
from BioConductor to store the modeling results and parameters. We extend the 
[SummarizedExperiment](http://bioconductor.org/packages/release/bioc/html/SummarizedExperiment.html) 
S4 class to create a new class RacipeSE to store and access the circuit/network,
simulated gene expressions, parameters, intial conditions and other meta 
information. As for any S4 class objects, the slots should not be accessed 
directly and the getter and accessor functions should be used instead. 

## Configuration

When a `RacipeSE` object is constructed, a `SummarizedExperiment` object is 
constructed and the default configuration is added as metadata. The default
configuration is stored as a `configData` data object. These configuration 
parameters can also be passed as argumnets to `sracipeSimulate` function.
The getter/accessor function for configuration is `sracipeConfig`.


```{r }

suppressWarnings(suppressPackageStartupMessages(library(sRACIPE)))

racipe <- RacipeSE() # Construct an empty RacipeSE object
racipe  # View the object. Notice the config in metadata. 
sracipeConfig(racipe) # access the config
```

The config list includes other lists or vectors like `simParams`, `stochParams`, 
`hyperParams`, `options`, `thresholds` etc. The list `simParams` contains values
for parameters like the number of models (`numModels`), simulation time
(`simulationTime`), step size for simulations (`integrateStepSize`), 
when to start recording the gene expressions (`printStart`), 
time interval between recordings (`printInterval`), 
number of initial conditions (`nIC`), output precision (`outputPrecision`), 
tolerance for adaptive runge kutta method (`rkTolerance`), 
parametric variation (`paramRange`). The list `stochParams` contains the 
parameters for stochastic simulations like the number of noise levels to be 
simulated (`nNoise`), the ratio of subsequent noise levels 
(`noiseScalingFactor`), maximum noise (`initialNoise`), whether to use same 
noise for all genes or to scale it as per the median expression of the genes 
(`scaledNoise`), ratio of shot noise to additive noise (`shotNoise`). 
The list `hyperParams` contains the parameters like the minimum and maximum 
production and degration of the genes, fold change, hill coefficient etc. 
The list `options` includes logical values like annealing (`anneal`), 
scaling of noise (`scaledNoise`), generation of new initial conditions 
(`genIC`), parameters (`genParams`) and whether to integrate or not 
(`integrate`). 


Metadata information also includes `annotation`, `nInteraction`
(number of interactions in the circuit), `normalized` 
(whether the data is normalized or not), data analysis lists like `pca`, `umap`,
cluster assignment of the models etc. These are added to the object in subsequent 
steps in the pipeline.


## Circuit
`SummarizedExperiment` slot `rowData` is used to store the circuit topology. 
It is a square matrix with dimension equal to the number of genes in the circuit. 
The values of the matrix represent the type of interaction in the gene pair given by row and column. 1 represents TF activation, 2 TF inhibition, 3 activation by degradation inhibition, 4 inhibition
by degradation activation, 5 signaling activation, 6 signaling inihibition,
and 0 for no interaction. The `sracipeCircuit` getter and 
accessor should be used to access this information. The `annotation` getter and 
setter set/get the name. Setting the circuit also adds `nInteractions` and 
`annotation` to the RacipeSE object.

```{r }
data("demoCircuit")
sracipeCircuit(racipe) <- demoCircuit
annotation(racipe) <- "demoCircuit"
annotation(racipe)
```


## Parameters and initial conditions
`SummarizedExperiment` slot `colData` contains the parameters and initial 
conditions for each model. Each gene in the circuit has two parameters, 
namely, its production rate and its degradation rate. Each interaction 
has three parameters, namely, threshold of activation, the hill coefficient, 
and the fold change. Each gene has one or more initial gene expression values 
as specified by nIC. The corresponding getter and setter are `sracipeParams`
and `sracipeIC`.


## Simulated gene expression
As such, sRACIPE can be used to simulate the trajectory for a single model 
with specific kinetic parameters
or a large large number of models with randomized parameters. 
These sRACIPE models can be considered as separate samples/cells differing from 
each other in terms of the parameters reflecting the cellular heterogeneity. 
The `SummarizedExperiment` slot `assays` is used for storing 
simulated gene expressions. 
The rows of these matrix-like elements correspond to various genes in the 
circuit and columns correspond to models. The first element of the `List` 
is used for unperturbed deterministic simulations. 
The subsequent elements are used for stochastic simulations at 
different noise levels and/or knockout simulations. The name for each assay 
contains the information on the noise level (for
stochastic simulations), knocked out gene (knockout simulations), 
or time (temporal simulations).
The `assay` or `assays` accessor and getters from the SummarizedExperiment class
should be used to access or modify these slots.

## Time Series
If one is interested in trajectories of a single model, then sRACIPE provides
an argument `timeSeries` in the `sracipeSimulate` funcion. If enabled, sRACIPE
will generate the trajectory for a single model and the corresponding plotting
functions will also show time series plots. In such case, the trajectories 
are stored in `metadata` instead of `assays`. The use of `sracipeGetTS` function 
is recommended in this case.

Usually, sRACIPE only records the steady states or the simulated gene expression
at the end of the simulation. But the model trajectories can also be recorded 
and analyzed using sRACIPE using the
variables `printInterval` and `printStart`. These variables define how 
often the simulated gene expressions are recorded. Thus, one can record multiple
time points during the model simulations. If intermediate time points 
are also recorded, then the first element will contain the last (steady-state)
solutions and the other slots will contain the values at the intermediate time 
points. For more details on temporal simulations of large number of models, 
please refer to the corresponding article.







